---
title: "MA615 Final Project"
author: "Yin Xu"
date: "2022-12-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE, message=FALSE}
library(readr)
library(tidyr)
library(dplyr)
library(ggradar)
library(tidyverse)
library(scales)
library(showtext)
library(ggplot2)
library(grid)
library(tidyverse)
library(shadowtext)
```

# MBTA Data EDA

This report includes the data from MBTA in last 12 months with all categories of transit (https://cdn.mbta.com/archive/archived_feeds.txt), and I picked one week from each month. The dataset is large, to run code faster and drop the useless data, I will do the data cleaning first.

In this report, I'm gonna graph the distribution of the `stops`. Besides, I found that typicality is worthy of analysis.
According to the reference(https://github.com/mbta/gtfs-documentation/blob/master/reference/gtfs.md#calendar_attributestxt), in this dataset, current valid values are:
0 (or empty): Not defined; 1: Typical service with perhaps minor modifications; 2: Extra service supplements typical schedules; 3: Reduced holiday service is provided by typical Saturday or Sunday schedule; 4: Major changes in service due to a planned disruption, such as construction; 5: Major reductions in service for weather events or other atypical situations.



## Read Data and Data Cleaning

```{r warning=FALSE, message=FALSE, include=FALSE}
Oct_22 <- read.csv("stop_times 22Oct.txt")
Sep_22 <- read.csv("stop_times 22Sep.txt")
Aug_22 <- read.csv("stop_times 22Aug.txt")
Jul_22 <- read.csv("stop_times 22Jul.txt")
Jun_22 <- read.csv("stop_times 22Jun.txt")
May_22 <- read.csv("stop_times 22May.txt")
Apr_22 <- read.csv("stop_times 22Apr.txt")
Mar_22 <- read.csv("stop_times 22Mar.txt")
Feb_22 <- read.csv("stop_times 22Feb.txt")
Jan_22 <- read.csv("stop_times 22Jan.txt")
Dec_21 <- read.csv("stop_times 21Dec.txt")
Nov_21 <- read.csv("stop_times 21Nov.txt")

stop_time_data <- rbind(Nov_21, Dec_21, Jan_22, Feb_22, Mar_22, Apr_22, May_22, Jun_22, Jul_22, Aug_22,
                        Sep_22, Oct_22)
```

```{r warning=FALSE, message=FALSE, echo=FALSE}
ca_22Oct <- read.csv("ca_22Oct.txt") %>% drop_na()
ca_22Sep <- read.csv("ca_22Sep.txt") %>% drop_na()
ca_22Aug <- read.csv("ca_22Aug.txt") %>% drop_na()
ca_22Jul <- read.csv("ca_22Jul.txt") %>% drop_na()
ca_22Jun <- read.csv("ca_22Jun.txt") %>% drop_na()
ca_22May <- read.csv("ca_22May.txt") %>% drop_na()
ca_22Apr <- read.csv("ca_22Apr.txt") %>% drop_na()
ca_22Mar <- read.csv("ca_22Mar.txt") %>% drop_na()
ca_22Feb <- read.csv("ca_22Feb.txt") %>% drop_na()
ca_22Jan <- read.csv("ca_22Jan.txt") %>% drop_na()
ca_21Dec <- read.csv("ca_21Dec.txt") %>% drop_na()
ca_21Nov <- read.csv("ca_21Nov.txt") %>% drop_na()

ca_data <- rbind(ca_21Nov, ca_21Dec, ca_22Jan, ca_22Feb, ca_22Mar, ca_22Apr, ca_22May, ca_22Jun,
                 ca_22Jul, ca_22Aug, ca_22Sep, ca_22Oct)
```


```{r, echo=FALSE}
stops <- read.csv("stops.txt")

df <- stops %>% inner_join(stop_time_data,by=c('stop_id'))
```

```{r warning=FALSE, echo=FALSE}
stoptime_new <- stop_time_data[,c(4,10)]
new_data <- stoptime_new %>%
  inner_join(stops, by = c('stop_id'))
new_data <- new_data[,c(1,2,4,10)]

newdata_df <- new_data %>% 
  count(zone_id) 
```

## Count stops

```{r}
newdata_df %>%
  ggplot( aes(x=zone_id, y=n)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    theme_bw()
```
The graph shows that bus has the most stops.



## Season vs Typicality
```{r}
#Clean Data
new_cadata <- ca_data %>%
  count(rating_description, service_schedule_typicality, sort = TRUE) %>%
  pivot_wider(names_from = service_schedule_typicality, values_from = n) 
new_cadata[is.na(new_cadata)] <- 0

new_cadata <- new_cadata %>%
  mutate(t_1 = `1`/(`1`+`2`+`3`+`4`+`5`),
         t_2 = `2`/(`1`+`2`+`3`+`4`+`5`),
         t_3 = `3`/(`1`+`2`+`3`+`4`+`5`),
         t_4 = `4`/(`1`+`2`+`3`+`4`+`5`),
         t_5 = `5`/(`1`+`2`+`3`+`4`+`5`)) 
new_cadata <- new_cadata[,c(1,7:11)]
```

```{r warning=FALSE, message=FALSE}
library(ggradar)
library(palmerpenguins)
library(tidyverse)
library(scales)
library(showtext)

font_add_google("Roboto", "roboto")
showtext_auto()

#plot1
#code citation: gg gallery
plot_1 <- new_cadata %>%
  ggradar(
    font.radar = "roboto",
    grid.label.size = 7,  
    axis.label.size = 7, 
    group.point.size = 3 
  )+
  labs(title = "Season vs Typicality") + 
  theme(
      plot.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4"),
      panel.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4"),
      plot.title.position = "plot",
      plot.title = element_text(
        family = "lobstertwo", 
        size = 20,
        face = "bold", 
        color = "#2a475e"
    )
  )
plot_1
```

According to this graph, we can know that in most seasons, the most typical service reported is typical service with perhaps minor modifications, but in summer/fall, major changes in service due to a planned disruption becomes the most, such as construction. Therefore, it can be concluded as MBTA generally tend to be constructed in summer/fall.


## Days vs Typicality
```{r}
#Clean Data
new_data_days <- ca_data %>%
  count(service_schedule_type, service_schedule_typicality, sort = TRUE) %>%
  pivot_wider(names_from = service_schedule_typicality, values_from = n) 
new_data_days[is.na(new_data_days)] <- 0

new_data_days <- new_data_days %>%
  mutate(t_1 = `1`/(`1`+`2`+`3`+`4`+`5`),
         t_2 = `2`/(`1`+`2`+`3`+`4`+`5`),
         t_3 = `3`/(`1`+`2`+`3`+`4`+`5`),
         t_4 = `4`/(`1`+`2`+`3`+`4`+`5`),
         t_5 = `5`/(`1`+`2`+`3`+`4`+`5`))
new_data_days <- new_data_days[,c(1,7:11)]
```


```{r warning=FALSE, message=FALSE}
plot_2 <- new_data_days %>%
  ggradar(
    font.radar = "roboto",
    grid.label.size = 7,
    axis.label.size = 7,
    group.point.size = 3) +
  labs(title = "Days vs Typicality") + 
  theme(
      plot.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4"),
      panel.background = element_rect(fill = "#fbf9f4", color = "#fbf9f4"),
      plot.title.position = "plot",
      plot.title = element_text(
        family = "lobstertwo", 
        size = 20,
        face = "bold", 
        color = "#2a475e"
    )
  )

plot_2
```

This graph shows that weekdays or weekends do not influence the typical service that much, weekdays have the less type4(Major changes in service due to a planned disruption, such as construction), MBTA might consider that the construction cannot influence the traffic in weekdays, so this kind of issue might be moved to saturdays and sundays.







